Komols Complexity, Coresets, and Sketches in Machine Learning

### 2019-01-08 komlos.tex

\begin{abstract}
\el{I need to rewrite the abstract}
This paper defines the Komlos complexity for families of functions which is intimately related to its discrepancy properties.
We argue that this measure of complexity captures properties like minimal coreset sizes for different problems in machine learning. We show that for many well known problems the Komlos complexity is asymptotically lower than their Rademacher complexity. This proves the existence smaller coresets than those achievable by random sampling. 
New results for coresets for kernel density estimation, matrix covariance approximation, and logistic regression are presented both as examples of for providing mechanisms to compute the Komlos complexity of other classes.
For all of the above we match the state of the art or improve on it.
This makes an explicit connection between low discrepancy, streaming algorithms, coresets, and learnability. 
%We also asymptotically improve on the generic merge-reduce framework for creating mergeable sketches for all the above problems. 
\end{abstract}

@@

\subsection{General Framework for Sketching}

The high level idea is the following. 
Consider a stream of the data points $x_i$. 
We maintain a buffer that can hold up to $m$ points.
Once collecting $m$ data points it performs a \emph{compact} operation that outputs at most $m/2$ data points of weight $2$ representing the original $m$ inputs. The compact operation partitions the $m$ points into two sets whose corresponding sums are very close. 
This is made formal with the following definitions:

 %
In what follows, we prove the following.
\begin{theorem} \label{thm:streaming}
For any function $f$ with a corresponding Komlos Complexity $K_m = O(c/m)$ there exists an unbiased fully-mergeable streaming coreset algorithms of size 
$$O\left(c\log^2\left(\min(\eps n/c, \log(n/\delta)) \right)/\eps\right)$$ 
whose error is at most $\eps n$ and failure probability at most $\delta$. For $\delta=0$ this results in a deterministic algorithm. The algorithm must be given $\eps, \delta$ as input and does not require a priori knowledge of the stream length $n$.
\end{theorem}

\begin{theorem} \label{thm:streaming2}
For any function $f$ with a corresponding Komlos Complexity of $K_m = O(c/\sqrt{m})$ there exists an unbiased fully-mergeable streaming coreset algorithms of size 
$$O\left(c\log^3\left(\min(\eps^2 n/c, \log(n/\delta)) \right)/\eps^2\right)$$ 
whose error is at most $\eps n$ and failure probability at most $\delta$. For $\delta=0$ this results in a deterministic algorithm. The algorithm must be given $\eps, \delta$ as input and does not require a priori knowledge of the stream length $n$.
\end{theorem}
%
%\begin{fact}
%For any function $f$ with a corresponding discrepancy of $\Delta_k = O(\sqrt{k})$ there exists an unbiased fully-mergeable streaming coreset algorithms of size 
%$O(\log^2(\log(1/\delta))/\eps^2)$ whose error is at most $\eps n$ and failure probability at most $\delta$.
%\end{fact}
%
%\begin{fact}
%For any function $f$ with a corresponding discrepancy of $\Delta_k = O(1)$ there exists a fully-mergeable deterministic streaming coreset algorithms of size 
%$O(\log^2(n)/\eps)$ whose error is at most $\eps n$.
%\end{fact}
%\begin{fact}
%For any function $f$ with a corresponding discrepancy of $\Delta_k = O(\sqrt{k})$ there exists a fully-mergeable deterministic streaming coreset algorithms of size 
%$O(\log^2(n)/\eps^2)$ whose error is at most $\eps n$.
%\end{fact}

@@

From section \ref{} we know that the Komlos complexity of the Gaussian kernel is $K_m = O(\sqrt{d}/m)$. Here, we show that for any positive kernel $K_m = O(1/\sqrt{m})$. This bound is better when $d > m$. More importantly though, there is a very simple, intuitive, and deterministic algorithm for computing the signs $\sigma$. 
Given a collection of data points $X = x_1,\ldots, x_n$ in $\R^d$ the density function $f: \R^d \rightarrow \R$ of a point $q$ is defined as 
$$ f(q) = \sum_{i=1}^{n} K(x_i,q) $$
Here, $K$ is a \emph{positive definite kernel} function, typically based on the distance between $x,y$. The most frequent examples include
$$ K(x,q) = \exp(- \|x-y\|_2^2/\lambda^2)\;\;\; K(x,q) = \exp(- \|x-q\|/\lambda) \; \mbox{and}\;\;\; K(x,y) = (1+\|x-q\|/_2^2/\lambda^2)^{-1}$$
where $\lambda$ is a scaling parameter. For simplicity, we assume that $K(x,x) \leq 1$ for all data points. Notice that for any kernel based on distance we have $K(x,x)=1$ exactly for all $x \in \R^d$.

Using our framework, we need to bound $K_m = \frac{1}{m}\sum_{i=1}^m \sigma_i K(x_i,q)$. For any kernel $K$ there exist a mapping $\phi: \R^d \to {\cal V}$ to an inner product space $\cal V$ such that 
$$ K(x,q) = \ip{\phi(x), \phi(q)} $$
Using this function $\phi$ our objective function becomes
\[
\sum_{i=1}^m \sigma_i K(x_i,q) = \sum_{i=1}^m \sigma_i \ip{\phi(x_i), \phi(q)} =  \ip{ \sum_{i=1}^m \sigma_i \phi(x_i), \phi(q)} \leq  \|\phi(q)\| \cdot \left\|  \sum_{i=1}^m \sigma_i \phi(x_i) \right\| %\leq \left\|  \sum_{i=1}^m \sigma_i \phi(x_i) \right\| 
\]
This upper bound allows us to find values for $\sigma_i$ which bound $K_n$ using a simple algorithm.
Set $\sigma_1 = 1$. Then for $i=2,\ldots,m$ set $\sigma_i = -\operatorname{sign} (\sum_{j=1}^{i-1}\sigma_j  K(x_j, x_i))$.
We will show by induction that $\| \sum_{j=1}^i \sigma_j \phi(x_j) \|^2 \le \sum_{j=1}^i \|\phi(x_j)\|^2$ for all $i$.
This is trivially true for $i=1$. 
Assume by induction that $\| \sum_{j=1}^{i-1} \sigma_j \phi(x_j)\|^2 \le \sum_{j=1}^{i-1} \|\phi(x_j)\|^2$.

@@

Using the framework above this provides a deterministic coreset construction for kernel density estimation of size $O(1/\eps^2)$ such that $\forall \;q\;\; |\tilde f(q) - f(q)| \le \eps n$. This matches, and greatly simplifies, the results achieved by \cite{DBLP:conf/soda/PhillipsT18} and \cite{DBLP:journals/corr/abs-1802-01751}.

>>>>

\begin{abstract}
This paper defines the Komlos complexity for families of functions and conjectures that it governs minimal coreset sizes for different problems in machine learning. Our result resolves a long standing open question regarding minimal cardinality coresets for kernel density estimation. It provides improved coreset for matrix covariance approximation, logistic regression, and any analytic function of the the dot product. The results leverage new tools and algorithms from discrepancy theory and provide tools for improving coreset sizes for other problems. Independently of the above, we show that any problem with low Komlos complexity admits a small merge-able streaming sketch. 
The result is an exponential improvement over the widely known and simple merge-and-reduce trick.
This paper makes an explicit connection between low discrepancy, coresets, learnability, and streaming algorithms. 
\end{abstract}

@@

\subsection{General Framework for Sketching}
We claim that low Komlos complexity implies concise streaming mergeable coresets. 
This assumes the existence of an algorithm for finding low discrepancy signs $\sigma$.
A trivial way to obtain such a construction is known as the merge-reduce trick \el{\cite{??} Har Peled?}.
\el{What do we get from merge-reduce here? $\log(n)/\eps^2$ or $\log^2(n)/\eps^2$? Is it like MRL?)}
The following theorems claim that a better construction is also possible.

%
%of function family $\F$ is low, 
%and one has an 
%
%The high level idea is the following. 
%Consider a stream of the data points $x_i$. 
%We maintain a buffer that can hold up to $m$ points.
%Once collecting $m$ data points it performs a \emph{compact} operation that outputs at most $m/2$ data points of weight $2$ representing the original $m$ inputs. The compact operation partitions the $m$ points into two sets whose corresponding sums are very close. 
%This is made formal with the following definitions: 
%
\begin{theorem} \label{thm:streaming}
For any function $f$ with a corresponding Komlos Complexity $K_m = O(c/m)$ there exists an fully-mergeable streaming coreset algorithms of size 
$$O\left(c\log^2\left(\min(\eps n/c, \log(n/\delta)) \right)/\eps\right)$$ 
whose error is at most $\eps n$ and failure probability at most $\delta$. For $\delta=0$ this results in a deterministic algorithm. The algorithm must be given $\eps, \delta$ as input and does not require a priori knowledge of the stream length $n$.
\end{theorem}

\begin{theorem} \label{thm:streaming2}
For any function $f$ with a corresponding Komlos Complexity of $K_m = O(c/\sqrt{m})$ there exists an fully-mergeable streaming coreset algorithms of size 
$$O\left(c\log^3\left(\min(\eps^2 n/c, \log(n/\delta)) \right)/\eps^2\right)$$ 
whose error is at most $\eps n$ and failure probability at most $\delta$. For $\delta=0$ this results in a deterministic algorithm. The algorithm must be given $\eps, \delta$ as input and does not require a priori knowledge of the stream length $n$.
\end{theorem}

The proofs of the above two theorems are quite technical and not very illuminating. Most of the ideas are already present in the recent breakthrough work on streaming quantiles \cite{DBLP:conf/focs/KarninLL16}. 
%
%\begin{fact}
%For any function $f$ with a corresponding discrepancy of $\Delta_k = O(\sqrt{k})$ there exists an unbiased fully-mergeable streaming coreset algorithms of size 
%$O(\log^2(\log(1/\delta))/\eps^2)$ whose error is at most $\eps n$ and failure probability at most $\delta$.
%\end{fact}
%
%\begin{fact}
%For any function $f$ with a corresponding discrepancy of $\Delta_k = O(1)$ there exists a fully-mergeable deterministic streaming coreset algorithms of size 
%$O(\log^2(n)/\eps)$ whose error is at most $\eps n$.
%\end{fact}
%\begin{fact}
%For any function $f$ with a corresponding discrepancy of $\Delta_k = O(\sqrt{k})$ there exists a fully-mergeable deterministic streaming coreset algorithms of size 
%$O(\log^2(n)/\eps^2)$ whose error is at most $\eps n$.
%\end{fact}

@@

From section \ref{} we know that the Komlos complexity of the Gaussian kernel is $K_m = O(\sqrt{d}/m)$. Here, we show that for any positive kernel $K_m = O(1/\sqrt{m})$. This bound is better when $d > m$. More importantly though, there is a very simple, intuitive, and deterministic algorithm for computing the signs $\sigma$. 
Given a collection of data points $X = x_1,\ldots, x_n$ in $\R^d$ the density function $F: \R^d \rightarrow \R$ of a point $q$ is defined as 
$$ F(q) = \sum_{i=1}^{n} K(x_i,q) $$
Here, $K$ is a \emph{positive definite kernel} function, typically based on the distance between $x,y$. The most frequent examples include
$$ K(x,q) = \exp(- \|x-y\|_2^2/\lambda^2)\;\;\; K(x,q) = \exp(- \|x-q\|/\lambda) \; \mbox{and}\;\;\; K(x,y) = (1+\|x-q\|/_2^2/\lambda^2)^{-1}$$
where $\lambda$ is a scaling parameter. For simplicity, we assume that $K(x,x) \leq 1$ for all data points. Notice that for any kernel based on distance we have $K(x,x)=1$ exactly for all $x \in \R^d$.

Using our framework, we need to bound $K_m = \frac{1}{m}\sum_{i=1}^m \sigma_i K(x_i,q)$. For any kernel $K$ there exist a mapping $\phi: \R^d \to {\cal V}$ to an inner product space $\cal V$ such that 
$$ K(x,q) = \ip{\phi(x), \phi(q)} $$
Using this function $\phi$ our objective function becomes
\[
\sum_{i=1}^m \sigma_i K(x_i,q) = \sum_{i=1}^m \sigma_i \ip{\phi(x_i), \phi(q)} =  \ip{ \sum_{i=1}^m \sigma_i \phi(x_i), \phi(q)} \leq  \|\phi(q)\| \cdot \left\|  \sum_{i=1}^m \sigma_i \phi(x_i) \right\| %\leq \left\|  \sum_{i=1}^m \sigma_i \phi(x_i) \right\| 
\]
This upper bound allows us to find values for $\sigma_i$ which bound $K_m$ using a simple algorithm.
Set $\sigma_1 = 1$. Then for $i=2,\ldots,m$ set $\sigma_i = -\operatorname{sign} (\sum_{j=1}^{i-1}\sigma_j  K(x_j, x_i))$.
We will show by induction that $\| \sum_{j=1}^i \sigma_j \phi(x_j) \|^2 \le \sum_{j=1}^i \|\phi(x_j)\|^2$ for all $i$.
This is trivially true for $i=1$. 
Assume by induction that $\| \sum_{j=1}^{i-1} \sigma_j \phi(x_j)\|^2 \le \sum_{j=1}^{i-1} \|\phi(x_j)\|^2$.

@@

Using the framework above this provides a deterministic coreset construction for any positive kernel density estimation of size $O(1/\eps^2)$ such that $\forall \;q\;\; |\tilde F(q) - F(q)| \le \eps n$. This matches, and greatly simplifies, the results achieved by \cite{DBLP:conf/soda/PhillipsT18} and \cite{DBLP:journals/corr/abs-1802-01751}.
